Unrecognized States: A Theory of Self-Determination and Foreign Influence

### 2015-11-17 draft.tex
A few typos

The first literature to address unrecognized states directly was grounded in comparative politics, and a robust area-studies literature exists around each of the current cases of unrecognized statehood.\footnote{Lynch (2004), King (2001), Stanislawski (2008), and Bakke et al. (2013) provide notable treatments of the Former Soviet cases, and a pair of edited volumes by Bahcheli, Bartmann, and Srebrnik (2004) and Kingston and Spears (2004) each compile broader sets of case studies.} More recent literature has addressed wider ranges of cases and made important conceptual progress identifying patterns and commonalities across cases (e.g. Pegg 1998; Kolst\o \ 2006; Geldenhuys 2009; Caspersen and Stansfield 2011; Caspersen 2013). Bakke (2011) summarizes four explanations for the persistence of unrecognized states, each of which plays a role in the theory we develop here. First, economic and political elites benefit from the status quo (e.g. King 2001). Second, the secessionists have sufficient military capacity to make military reconquest costly (e.g. Kolst\o \ 2006). Third, secessionists enjoy external support from a patron (e.g. Lynch 2002; Stanislawski 2008. Lastly, the governments of unrecognized states are internally legitimate, enjoying support from residents of the unrecognized state (e.g. Lynch 2002, Kolst\o \ 2006).

@@

Although we are not aware of any formal models of unrecognized statehood there is a growing economics literature that employs game theoretic models of conflict. An early model of territorial disputes in which the divisibility of the contested territory is an important parameter is Grossman (2004). In Hirschleifer, Boldrin and Levin (2009), conflict can be avoided when there is common knowledge, a common rate of time preference, and players can make a series of small concessions. In our context, the contested territory is indivisible; this implies that peace would require very large concessions so that this pathway of using small compromises to avoid conflict is not available.

@@

A major innovation of our model is the introduction of international actors. While the role of outside actors in determining the duration and outcome of civil conflicts is well documented (e.g., Elbadawi and Sambanis 2000; Regan 2002; Balch-Lindsay, Enterline, and Joyce 2008), the role of these actors has not been incorporated into the modeling of these conflicts.\footnote{One exception is van Houten (1998), who models the patron state ("reference state'') as a player in ethnic conflicts but otherwise takes an approach quite different from ours.} This is true even of work that addresses the role of outside actors as potential third-party enforcers (Walter 1997, 2002).

@@

We construct a model with four players: the secessionist elite ($s$); the central government of the home state ($g$) from which $s$ is attempting to secede; and two other players, the patron ($p$) and the international community ($c$). The latter two are outside actors---states, groups of states and/or individuals acting in concert---that have interests in the outcome of the attempted secession.

@@

At the start of each period, the patron, $p$, and player $c$ may invest resources to affect the payoffs of $s$ and $g$ in the stage game. For example, if $p$ expends resources $R_{pt}$ to increase the secessionists status quo payoffs, the beginning-of-period payoff $q_{st}$ becomes end-of-period payoff $Q_{st} = q_{st} + R_{pt}$. We can write analogous expressions for $q_{gt}, \ l_{st},\ l_{gt},\ w_{st}$, and $w_{st}$, and the international community can also make investments to increase whichever of these payoffs they like. $B_{pt}$ denotes the total that the patron can devote to interceding in the conflict; we assume that player $c$, which is often a large coalition of states, does not have a binding budget constraint.

@@

In equilibrium, the outside actors' investments are paired: the patron's investments in the secessionists' status quo payoffs deter player $c$'s investments in the secessionists' payoffs from ceding. Likewise, investments by player $c$ in the government's status quo payoffs deter the patron from investing in the government's payoffs from ceding. Investments by player $c$ in the secessionists' status quo payoffs deter the patron from investing in the secessionists' payoffs from winning the conflict via fighting. In each case, the potential  investments are such that the largest continuation payoffs in the game between the secessionists and the government those from playing status quo.

@@

In Section~\ref{sec:mainproof}, we considered the outside actors' abilities to make investments to increase the various payoffs of the home state government and the secessionists. The international community, in particular, often employs another option by joining the home state in enforcing economic sanctions against the unrecognized state, an action that \emph{reduces} the secessionists' payoffs from the status quo (i.e. sanctions are equivalent to $R_{ct} < 0$). Note that this may be particularly effective if $c$ is a large coalition of states acting in concert.

@@

\section{Conclusion}
In this paper we establish unrecognized states as an outcome of interest for social scientists and provide a unified framework for analyzing that outcome and its alternatives. Current events, particularly, Russia's military and economic support of secessionist rebels in Eastern Ukraine, suggest that the phenomenon of unrecognized statehood will not fade from relevance soon. While the Ukrainian situation remains fluid, as of October 2015, the status quo equilibrium we model is a plausible future. The equilibrium could emerge as follows: with Russian support, the secessionists solidify \emph{de facto} control of parts of Eastern Ukraine, but risk of direct confrontation with the West is sufficiently high to prevent Russian annexation of Eastern Ukraine.\footnote{Annexation of Crimea was achieved at a reasonable cost to Russia. The costs of annexation Eastern Ukraine would likely be substantially higher.} Further Russian economic aid keeps the status quo payoffs of the secessionists sufficiently high that they decline negotiated reunification. If necessary, contributions to the home state by U.S. and the EU (player $c$) strengthen the home state enough to prevent the secessionists from overthrowing the Ukrainian government and forcing recognition. In twenty years, Eastern Ukraine may look a good deal like South Ossetia does today.

>>>>

The first literature to address unrecognized states directly was grounded in comparative politics, and a robust area-studies literature exists around each of the current cases of unrecognized statehood.\footnote{Lynch (2004), King (2001), Stanislawski (2008), and Bakke et al. (2013) provide notable treatments of the Former Soviet cases, and a pair of edited volumes by Bahcheli, Bartmann, and Srebrnik (2004) and Kingston and Spears (2004) each compile broader sets of case studies.} More recent literature has addressed wider ranges of cases and made important conceptual progress identifying patterns and commonalities across cases (e.g. Pegg 1998; Kolst\o \ 2006; Geldenhuys 2009; Caspersen and Stansfield 2011; Caspersen 2013). Bakke (2011) summarizes four explanations for the persistence of unrecognized states, each of which plays a role in the theory we develop here. First, economic and political elites benefit from the status quo (e.g. King 2001). Second, the secessionists have sufficient military capacity to make military reconquest costly (e.g. Kolst\o \ 2006). Third, secessionists enjoy external support from a patron (e.g. Lynch 2002; Stanislawski 2008). Lastly, the governments of unrecognized states are internally legitimate, enjoying support from residents of the unrecognized state (e.g. Lynch 2002, Kolst\o \ 2006).

@@

Although we are not aware of any formal models of unrecognized statehood there is a growing economics literature that employs game theoretic models of conflict. Grossman (2004) offers an early model of territorial disputes in which the divisibility of the contested territory is an important parameter. In Hirschleifer, Boldrin and Levin (2009), conflict can be avoided when there is common knowledge, a common rate of time preference, and players can make a series of small concessions. In our context, the contested territory is indivisible; this implies that peace would require very large concessions so that this pathway of using small compromises to avoid conflict is not available.

@@

One innovation of our model is the introduction of international actors. While the role of outside actors in determining the duration and outcome of civil conflicts is well documented (e.g., Elbadawi and Sambanis 2000; Regan 2002; Balch-Lindsay, Enterline, and Joyce 2008), the role of these actors has not been incorporated into the modeling of these conflicts.\footnote{One exception is van Houten (1998), who models the patron state ("reference state'') as a player in ethnic conflicts but otherwise takes an approach quite different from ours.} This is true even of work that addresses the role of outside actors as potential third-party enforcers (Walter 1997, 2002).

@@

We construct a model with four players: the secessionist elite ($s$), which seeks recognized independence; the central government of the home state ($g$) from which $s$ is attempting to secede, which seeks reunification; and two other players, the patron ($p$) and the international community ($c$). The latter two are outside actors---states, groups of states and/or individuals acting in concert---that have interests in the outcome of the attempted secession.

@@

At the start of each period, the patron, $p$, and player $c$ may invest resources to affect the payoffs of $s$ and $g$ in the stage game. For example, if $p$ expends resources $R_{pt}$ to increase the secessionists' status quo payoffs, the beginning-of-period payoff $q_{st}$ becomes end-of-period payoff $Q_{st} = q_{st} + R_{pt}$. We can write analogous expressions for $q_{gt}, \ l_{st},\ l_{gt},\ w_{st}$, and $w_{st}$, and the international community can also make investments to increase whichever of these payoffs they like. $B_{pt}$ denotes the total that the patron can devote to interceding in the conflict; we assume that player $c$, which is often a large coalition of states, does not have a binding budget constraint.

@@

In equilibrium, the outside actors' investments are paired: the patron's investments in the secessionists' status quo payoffs deter player $c$'s investments in the secessionists' payoffs from ceding. Likewise, investments by player $c$ in the government's status quo payoffs deter the patron from investing in the government's payoffs from ceding. Investments by player $c$ in the secessionists' status quo payoffs deter the patron from investing in the secessionists' payoffs from winning the conflict via fighting. In each case, the potential  investments are such that the largest continuation payoffs in the game between the secessionists and the government are those from playing status quo.

@@

In Section~\ref{sec:mainproof}, we considered the outside actors' abilities to make investments to increase the various payoffs of the home state government and the secessionists. Player $c$, in particular, often employs another option by joining the home state in enforcing economic sanctions against the unrecognized state, an action that \emph{reduces} the secessionists' payoffs from the status quo (i.e. sanctions are equivalent to $R_{ct} < 0$). Note that this may be particularly effective if $c$ is a large coalition of states acting in concert.

@@

\section{Conclusion}
In this paper we establish unrecognized states as an outcome of interest for social scientists and provide a unified framework for analyzing that outcome and its alternatives. Current events, particularly Russia's military and economic support of secessionist rebels in Eastern Ukraine, suggest that the phenomenon of unrecognized statehood will not fade from relevance soon. While the Ukrainian situation remains fluid, as of October 2015, the status quo equilibrium we model is a plausible future. The equilibrium could emerge as follows: with Russian support, the secessionists solidify \emph{de facto} control of parts of Eastern Ukraine, but risk of direct confrontation with the West is sufficiently high to prevent Russian annexation of Eastern Ukraine.\footnote{Annexation of Crimea was achieved at a reasonable cost to Russia. The costs of annexation Eastern Ukraine would likely be substantially higher.} Further Russian economic aid keeps the status quo payoffs of the secessionists sufficiently high that they decline negotiated reunification. If necessary, contributions to the home state by U.S. and the EU (player $c$) strengthen the home state enough to prevent the secessionists from overthrowing the Ukrainian government and forcing recognition. In twenty years, Eastern Ukraine may look a good deal like South Ossetia does today.
